---
title: "An R package for change point detection: dSTEM"
author: Zhibing He
date: June 20, 2023
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r echo=FALSE}
library(dSTEM)
library(latex2exp)
set.seed(2021)
```
## dSTEM: differential Smoothing and TEsting of Maxima/Minima

A new generic and low-computational approach for multiple change points detection in linear models. The method is designed for the data sequence modeled as piecewise linear signal plus correlated random noise (a stationary Gaussian process) <https://doi.org/10.1214/20-EJS1751>.
The research paper for change point detection of general piecewise linear model will be available soon...

## Toy Example of the Method

The core idea of our method is to transform the change points into local maxima/minima of the differential smoothed data. By performing 
multiple testing on the local maxima/minima, the significant ones are detected as change points.

For example, a continuous change point will be transformed into a local maximum/minimum in the second derivative of the smoothed data, 
while a noncontinuous change point will be transformed into a local maximum/minimum in the first derivative of the smoothed data.
A toy example is illustrated in
```{r, echo = FALSE}
k1 = 0.08
k2 = 0.15
gamma=8
v = 100 # change point location
c = 4 # bandwidth of smoothing kernel
d1mu = function(x) a/gamma*dnorm((v-x)/gamma)+(k1-k2)*pnorm((v-x)/gamma)+k2
d2mu = function(x) (a*(v-x)+(k2-k1)*gamma^2)/gamma^3*dnorm((v-x)/gamma)
#par(mfrow=c(3,3),mar=c(2,2.5,2,0.5))
par(mfrow=c(3,3),mar=c(4,4,2,1))
#### Type I change Point
a = 0
b = a - (k2-k1)*v
xmax = 200
ymax = k2*xmax+b
x = 1:xmax
curve(k1*x,0,floor(xmax/2),xlim = c(0,xmax),ylim=c(0,ymax+1),lwd=2,xlab="(a)",ylab="Scenario 1",main=TeX("$\\mu(t)$"))
curve(k2*x+b,floor(xmax/2),xmax,add=TRUE,lwd=2,xlab="",ylab="")
points(x=v,y=k1*v,pch=16)
points(x=v,y=k2*v+b,pch=1)
abline(v=v,col="red",lty=2)
y = d1mu(x)
plot(loess(y~x,span=0.6),type="l",lwd=2,xlab="(b)",ylab="",main=TeX("$\\mu_{\\gamma}^{\\prime}(t)$"))
abline(v=v,col="red",lty=2,lwd=1)
points(x = c(v-c*gamma,v+c*gamma),y=y[c(v-c*gamma,v+c*gamma)],pch =16 ,col="blue")
y = d2mu(x)
plot(loess(y~x,span=0.6),type="l",lwd=2,xlab="(c)",ylab="",main=TeX("$\\mu_{\\gamma}^{\\prime\\prime}(t)$"))
abline(v=v,col="red",lty=2,lwd=1)
points(x = c(v-c*gamma,v+c*gamma),y=y[c(v-c*gamma,v+c*gamma)],pch =16 ,col="blue")
#### Type II Change Point
a = 3
b = a - (k2-k1)*v
xmax = 200
ymax = k2*xmax+b
x = 1:xmax
curve(k1*x,0,floor(xmax/2),xlim = c(0,xmax),ylim=c(0,ymax+1),lwd=2,xlab="(d)",ylab="Scenario 2")
curve(k2*x+b,floor(xmax/2),xmax,add=TRUE,lwd=2,xlab="",ylab="")
points(x=v,y=k1*v,pch=16)
points(x=v,y=k2*v+b,pch=1)
abline(v=v,col="red",lty=2)
y = d1mu(x)
plot(loess(y~x,span=0.6),type="l",lwd=2,xlab="(e)",ylab="")
abline(v=v,col="red",lty=2,lwd=1)
points(x = c(v-c*gamma,v+c*gamma),y=y[c(v-c*gamma,v+c*gamma)],pch =16 ,col="blue")
y = d2mu(x)
plot(loess(y~x,span=0.6),type="l",lwd=2,xlab="(f)",ylab="")
abline(v=v,col="red",lty=2,lwd=1)
points(x = c(v-c*gamma,v+c*gamma),y=y[c(v-c*gamma,v+c*gamma)],pch =16 ,col="blue")
#### Mixture of Type I and Type II Change Points
a = 0
b = a - (k2-k1)*v
xmax1 = 200 ; xmax2 = 300
ymax = k2*xmax1+b
x = 1:xmax2
a3 = -3 ; k3 = 0 ; v3 = xmax1
b3 = a3 - (k3-k2)*v3 +b
curve(k1*x,0,floor(xmax1/2),xlim = c(0,xmax2),ylim=c(0,ymax+1),lwd=2,xlab="(g)",ylab="Scenario 3")
curve(k2*x+b,floor(xmax1/2),xmax1,add=TRUE,lwd=2,xlab="",ylab="")
points(x=v,y=k1*v,pch=16)
abline(v=v,col="red",lty=2)
curve(k3*x+b3,xmax1,xmax2,add=TRUE,lwd=2,xlab="",ylab="")
abline(v=v3,col="red",lty=2)
points(x=v3,y=k2*v3+b,pch=16)
points(x=v3,y=k3*v3+b3,pch=1)
#
y1 = d1mu(x[1:floor(xmax2/2)])
plot(loess(y1~x[1:floor(xmax2/2)],span=0.6),type="l",lwd=2,xlab="(h)",ylab="",xlim=c(0,xmax2),ylim=c(-0.10,0.15))
abline(v=v,col="red",lty=2,lwd=1)
points(x = c(v-c*gamma,v+c*gamma),y=y1[c(v-c*gamma,v+c*gamma)],pch =16 ,col="blue")
k1=k2;k2=k3;a=a3;v=v3
y2 = d1mu(x[floor(xmax2/2):xmax2])
lines(loess(y2~x[floor(xmax2/2):xmax2],span=0.6),lwd=2,xlab="",ylab="")
abline(v=v3,col="red",lty=2,lwd=1)
points(x = c(v3-c*gamma,v3+c*gamma),y=y2[c(v3-c*gamma,v3+c*gamma)-floor(xmax2/2)],pch =16 ,col="blue")
#
k1 = 0.08;k2 = 0.15;v = 100;a = 0
y1 = d2mu(x[1:floor(xmax2/2)])
plot(loess(y1~x[1:floor(xmax2/2)],span=0.6),type="l",lwd=2,xlab="(i)",ylab="",xlim=c(0,xmax2),ylim=c(-0.016,0.007))
abline(v=v,col="red",lty=2,lwd=1)
points(x = c(v-c*gamma,v+c*gamma),y=y1[c(v-c*gamma,v+c*gamma)],pch =16 ,col="blue")
k1=k2;k2=k3;a=a3;v=v3
y2 = d2mu(x[floor(xmax2/2):xmax2])
lines(loess(y2~x[floor(xmax2/2):xmax2],span=0.6),lwd=2,xlab="",ylab="")
abline(v=v3,col="red",lty=2,lwd=1)
points(x = c(v3-c*gamma,v3+c*gamma),y=y2[c(v3-c*gamma,v3+c*gamma)-floor(xmax2/2)],pch =16 ,col="blue")
```

```{r, echo = FALSE}
l = 1200
h = seq(150,by=150,length.out=6)
jump = rep(0,7)
beta1 = c(2,-1,2.5,-3,-0.2,2.5)/50
beta1 = c(beta1,-sum(beta1*(c(h[1],diff(h))))/(l-tail(h,1)))
signal = gen.signal(l,h,jump,beta1)
#signal = gen.signal(l,h,jump,beta1,rep=10)
noise = rnorm(length(signal),0,1)
data = (signal+noise)
#modeling
gamma = 20
dy = diff(smth.gau(data,gamma),na.rm=T)
ddy = diff(dy[!is.na(dy)])
model1_2 = dstem(data,type="I",gamma,alpha = 0.05)
##
par(mfrow = c(2,2),mar=c(4,4,2,2))
b = 8
##
col1 = "blue"; col2 = "darkorange"
plot(data,col="grey",xlab="",ylab="",bty="n")
title(xlab="t",ylab="y(t)",mgp=c(2.5,2.5,0),cex.lab=1.0)
lines(signal,lwd=2,col="red")
plot(ddy,type="l",lwd=2,xlab="",ylab="",bty="n")
title(xlab="t",ylab=latex2exp::TeX("$y''_{\\gamma}(t)"),mgp=c(2.5,2.5,0),cex.lab=1.0)
abline(h=0,lty=1,lwd=2,col="lightgrey")
peak = which.peaks(ddy); val = which.peaks(ddy,decreasing = T)
points(peak,ddy[peak],col=col1,pch=16,bg=col1)
points(val,ddy[val],col=col2,pch=16,bg=col2)
tpeak = model1_2$peak; tval = model1_2$vall
points(tpeak,ddy[tpeak],col=col1,pch=24,bg=col1)
points(tval,ddy[tval],col=col2,pch=25,bg=col2)
Tmax = h[which(h %in% which.peaks(signal,decreasing=T))]; Tmin = h[which(h %in% which.peaks(signal))]
for(i in 1:length(Tmax)) rect(Tmax[i]-b,0,Tmax[i]+b,1,col="cyan",lwd=1.5,density=0)
for(i in 1:length(Tmin)) rect(Tmin[i]-b,0,Tmin[i]+b,-1,col="pink",lwd=1.5,density=0)

#### Type II change points
l = 1200
h = seq(150,by=150,length.out=6)
jump = c(0,1.5,2,-2.2,1.8,2,-1.5)*2
beta1 = c(2,-1,1.5,-2.5,-0.2,2.5,-1)/50
signal = gen.signal(l,h,jump,beta1)
noise = rnorm(length(signal),0,1)
data = (signal+noise)
# modeling
gamma = 10
dy = diff(smth.gau(data,gamma),na.rm=T)
ddy = diff(dy[!is.na(dy)])
# estimated location and slope
model2_1 = dstem(data,type="II-linear",alpha=0.05)
##
col3="green"; col4 = "red"
plot(data,col="grey",xlab="",ylab="",bty="n")
title(xlab="t",ylab="y(t)",mgp=c(2.5,2.5,0),cex.lab=1.0)
#
segment = split(1:l,cut(1:l,breaks=c(1,h,l),include.lowest = T))
for(i in 1:length(segment)) lines(x=segment[[i]],y=signal[segment[[i]]],lwd=2,col="red")
#
plot(dy,type="l",lwd=2,xlab="",ylab="",bty="n")
title(xlab="t",ylab=latex2exp::TeX("$y'_{\\gamma}(t)"),mgp=c(2.5,2.5,0),cex.lab=1.0)
abline(h=0,lty=1,lwd=2,col="lightgrey")
peak = which.peaks(dy); val = which.peaks(dy,decreasing = T)
points(peak,dy[peak],col=col3,pch=16,bg=col3)
points(val,dy[val],col=col4,pch=16,bg=col4)
tpeak = model2_1$peak; tval = model2_1$vall
points(tpeak,dy[tpeak],col=col3,pch=24,bg=col3)
points(tval,dy[tval],col=col4,pch=25,bg=col4)
Tmin = h[c(3,6)]; Tmax = h[c(1,2,4,5)]
for(i in 1:length(Tmax)) rect(Tmax[i]-b,0,Tmax[i]+b,1,col="cyan",lwd=1.5,density=0)
for(i in 1:length(Tmin)) rect(Tmin[i]-b,0,Tmin[i]+b,-1,col="pink",lwd=1.5,density=0)
for(i in 1:length(segment)) lines(x=segment[[i]],y=rep(beta1[i],length(segment[[i]])),
                                  lwd=2,col="darkgrey")
##### Mixture of type I and type II
l = 1500
h = seq(150,by=150,length.out=8)
jump = c(0,0,2.5,0,3,0,0,2.8,-2.5)
beta1 = c(1,-1,1.5,-2,-0.2,1.2,-0.5,1.2,0)/50
signal = gen.signal(l,h,jump,beta1)
noise = rnorm(length(signal),0,1)
data = (signal+noise)
## Modeling
gamma = 25
dy = diff(smth.gau(data,gamma),na.rm=T)
ddy = diff(dy[!is.na(dy)])
model2 = cpTest(x=ddy,order=2,gamma=gamma,alpha=0.1)
breaks = est.pair(model2$vall,model2$peak,gamma)$cp
slope = est.slope(data,breaks)
model1 = dstem(data,type="mixture",gamma,alpha = 0.05)$type2
model2 = dstem(data,type="mixture",gamma,alpha = 0.05)$type1
##
par(mfrow = c(1,3),mar=c(4,4,2,2))
plot(data,col="grey",xlab="",ylab="",bty="n")
title(xlab="t",ylab="y(t)",mgp=c(2.5,2.5,0),cex.lab=1.0)
#
segment = split(1:l,cut(1:l,breaks=c(1,h[c(2,4,7,8)],l),include.lowest = T))
for(i in 1:length(segment)) lines(x=segment[[i]],y=signal[segment[[i]]],lwd=2,col="red")
#
b=15
plot(dy,type="l",lwd=2,xlab="",ylab="",bty="n")
title(xlab="t",ylab=latex2exp::TeX("$y'_{\\gamma}(t)"),mgp=c(2.5,2.5,0),cex.lab=1.0)
abline(h=0,lty=1,lwd=2,col="lightgrey")
peak = which.peaks(dy); val = which.peaks(dy,decreasing = T)
points(peak,dy[peak],col=col3,pch=16,bg=col3)
points(val,dy[val],col=col4,pch=16,bg=col4)
tpeak = model1$peak; tval = model1$vall
points(tpeak,dy[tpeak],col=col3,pch=24,bg=col3)
points(tval,dy[tval],col=col4,pch=25,bg=col4)
Tmin = h[8]; Tmax = h[c(2,4,7)]
for(i in 1:length(Tmax)) rect(Tmax[i]-b,0,Tmax[i]+b,1,col="cyan",lwd=1.5,density=0)
for(i in 1:length(Tmin)) rect(Tmin[i]-b,0,Tmin[i]+b,-1,col="pink",lwd=1.5,density=0)
#
segment = split(1:l,cut(1:l,breaks=c(1,h,l),include.lowest = T))
for(i in 1:length(segment)) lines(x=segment[[i]],y=rep(beta1[i],length(segment[[i]])),
                                 lwd=2,col="darkgrey")
#
plot(ddy,type="l",lwd=2,xlab="",ylab="",bty="n")
title(xlab="t",ylab=latex2exp::TeX("$y''_{\\gamma}(t)"),mgp=c(2.5,2.5,0),cex.lab=1.0)
abline(h=0,lty=1,lwd=2,col="lightgrey")
peak = which.peaks(ddy); val = which.peaks(ddy,decreasing = T)
points(peak,ddy[peak],col=col1,pch=16,bg=col1)
points(val,ddy[val],col=col2,pch=16,bg=col2)
tpeak = model2$peak; tval = model2$vall
points(tpeak,ddy[tpeak],col=col1,pch=24,bg=col1)
points(tval,ddy[tval],col=col2,pch=25,bg=col2)
Tmin = h[c(1,3,6)]; Tmax = h[5]
for(i in 1:length(Tmax)) rect(Tmax[i]-b,0,Tmax[i]+b,1,col="cyan",lwd=1.5,density=0)
for(i in 1:length(Tmin)) rect(Tmin[i]-b,0,Tmin[i]+b,-1,col="pink",lwd=1.5,density=0)
Tmax = c(model1$peak,model1$vall)
for(i in 1:length(Tmax)) rect(Tmax[i]-2*gamma,0,Tmax[i]+1.5*gamma,1,col="grey",lwd=1.5,density=10)
for(i in 1:length(Tmax)) rect(Tmax[i]-2*gamma,0,Tmax[i]+1.5*gamma,-1,col="grey",lwd=1.5,density=10)
```



## Usage

The main function is 'dstem()', which detect the change points in piecewise constant and piecewise linear signals.

```{r pressure, echo=TRUE}
l = 1200
h = seq(150,by=150,length.out=6)
jump = rep(0,7)
beta1 = c(2,-1,2.5,-3,-0.2,2.5)/50
beta1 = c(beta1,-sum(beta1*(c(h[1],diff(h))))/(l-tail(h,1)))
signal = gen.signal(l,h,jump,beta1)
noise = rnorm(length(signal),0,1)
gamma = 25
unlist(dstem(signal + noise,"I",gamma=gamma,alpha=0.05))

## piecewise constant
l = 1200
h = seq(150,by=150,length.out=6)
jump = c(0,1.5,2,2.2,1.8,2,1.5)
beta1 = rep(0,length(h)+1)
signal = gen.signal(l,h,jump,beta1)
noise = rnorm(length(signal),0,1)
gamma = 25
unlist(dstem(signal + noise, "II-step",gamma,alpha=0.05))

## piecewise linear with jump
l = 1200
h = seq(150,by=150,length.out=6)
jump = c(0,1.5,2,2.2,1.8,2,1.5)*3
beta1 = c(2,-1,2.5,-3,-0.2,2.5,-0.5)/50
signal = gen.signal(l=l,h=h,jump=jump,b1=beta1)
noise = rnorm(length(signal),0,1)
gamma = 25
unlist(dstem(signal + noise, "II-linear",gamma,alpha=0.05))
```

